---
title: "Homework 1"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
# install.packages("nycflights13")   # run once, then leave this commented out
library(nycflights13)
data(flights)
```

<!--
This skeleton exists so you spend your time on the analysis rather than on
YAML. It is deliberately thin: the headings are here, the answers are not, and
deciding what code belongs under each heading is part of what is being graded.

Before you submit: Session > Restart R, then Knit. If it does not knit from a
clean session, it will not knit on my machine either.

Every number you state in prose must come from inline R code -- like this:
`r nrow(flights)` rows -- and not be typed by hand.
-->

# Part 1 · The question (8 points)

*A S.M.A.R.T. question this file can answer, and one it cannot. No code.*

**A question this file can answer.**

**A question it cannot, and what specifically blocks it.**

# Part 2 · One variable, honestly (13 points)

## (a) Five numbers and the IQR
```{r}
summary(flights$arr_delay)
cat("IQR:", IQR(flights$arr_delay, na.rm = TRUE), "\n")
cat("SD:", sd(flights$arr_delay, na.rm = TRUE), "\n")
```



## (b) Mean against median, checked with a histogram
The mean (~6.9), is larger than the median (-5). This is indicative of a right-skewed distribution.
```{r}

mean <- mean(flights$arr_delay, na.rm = TRUE)
median <- median(flights$arr_delay, na.rm = TRUE)
cat("mean:", mean, "\n")
cat("median:", median, "\n")
cat("mean > median:", mean > median)

hist(flights$arr_delay,
     main = "Histogram of Arrival Delay",
     xlab = "Delay (minutes)")

abline(v = mean(flights$arr_delay, na.rm = TRUE), col = "red", lwd = 2, lty = 2)
abline(v = median(flights$arr_delay, na.rm = TRUE), col = "blue", lwd = 2, lty = 1)

legend("right",
       legend = c("Mean", "Median"),
       col = c("red", "blue"),
       lwd = 2,
       lty = c(2, 1))

```


## (c) The proportion arriving before the mean
Roughly, 67.7% of flights arrive earlier than the average flight delay time, which tells us that most of the flights arrive nearly on time if not early.
```{r}
# proportion of flights that arrive earlier than the mean arrival delay
cat("mean:", mean, "\n")
cat("flight delays < mean:",
  mean(mean(flights$arr_delay, na.rm = TRUE) > flights$arr_delay, na.rm = TRUE))

```


## (d) IQR against the standard deviation
The IQR/SD ratio is 0.69, this indicates a distribution with heavy tails.
```{r}
cat("IQR/SD:",
IQR(flights$arr_delay, na.rm = TRUE) / sd(flights$arr_delay, na.rm = TRUE))

```



# Part 3 · The flights that are not there (23 points)

## (a) How many, and what percentage

## (b) The two mechanisms

## (c) What `na.rm = TRUE` actually describes

*Your sentence starts "The average arrival delay of..."*

## (d) Is the deletion carrier-neutral?

*Answer with a number.*

## (e) What you would do about it

# Part 4 · A ranking that is not what it looks like (18 points)

## (a) Carriers by mean arrival delay

## (b) What should stop you in the top five

## (c) Which panel this is

## (d) A rule, applied

# Part 5 · A plot that misleads, and its honest twin (13 points)

## (a) The misleading plot

## (b) The honest plot

## (c) What changed, and which change did the work

# Part 6 · Who is missing (15 points)

*One paragraph, at most 150 words. Three mechanisms, each with a direction.*

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->