---
title: "Homework 1"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
# install.packages("nycflights13")   # run once, then leave this commented out
library(nycflights13)
library(tidyverse)
data(flights)
```

<!--
This skeleton exists so you spend your time on the analysis rather than on
YAML. It is deliberately thin: the headings are here, the answers are not, and
deciding what code belongs under each heading is part of what is being graded.

Before you submit: Session > Restart R, then Knit. If it does not knit from a
clean session, it will not knit on my machine either.

Every number you state in prose must come from inline R code -- like this:
`r nrow(flights)` rows -- and not be typed by hand.
-->

# Part 1 · The question (8 points)

*A S.M.A.R.T. question this file can answer, and one it cannot. No code.*

**A question this file can answer.**

**A question it cannot, and what specifically blocks it.**

# Part 2 · One variable, honestly (13 points)

## (2a) Five numbers and the IQR
```{r}
summary(flights$arr_delay)
cat("IQR:", IQR(flights$arr_delay, na.rm = TRUE), "\n")
cat("SD:", sd(flights$arr_delay, na.rm = TRUE), "\n")
```



## (2b) Mean against median, checked with a histogram
The mean (~6.9), is larger than the median (-5). This is indicative of a right-skewed distribution.
```{r}

mean <- mean(flights$arr_delay, na.rm = TRUE)
median <- median(flights$arr_delay, na.rm = TRUE)
cat("mean:", mean, "\n")
cat("median:", median, "\n")
cat("mean > median:", mean > median)

hist(flights$arr_delay,
     main = "Histogram of Arrival Delay",
     xlab = "Delay (minutes)")

abline(v = mean(flights$arr_delay, na.rm = TRUE), col = "red", lwd = 2, lty = 2)
abline(v = median(flights$arr_delay, na.rm = TRUE), col = "blue", lwd = 2, lty = 1)

legend("right",
       legend = c("Mean", "Median"),
       col = c("red", "blue"),
       lwd = 2,
       lty = c(2, 1))

```


## (2c) The proportion arriving before the mean
Roughly, 67.7% of flights arrive earlier than the average flight delay time, which tells us that most of the flights arrive nearly on time if not early.
```{r}
# proportion of flights that arrive earlier than the mean arrival delay
cat("mean:", mean, "\n")
cat("flight delays < mean:",
  mean(mean(flights$arr_delay, na.rm = TRUE) > flights$arr_delay, na.rm = TRUE))

```


## (2d) IQR against the standard deviation
The IQR/SD ratio is 0.69, this indicates a distribution with heavy tails.
```{r}
cat("IQR/SD:",
IQR(flights$arr_delay, na.rm = TRUE) / sd(flights$arr_delay, na.rm = TRUE))

```



# Part 3 · The flights that are not there (23 points)

## (3a) How many, and what percentage
There are 9430 rows that are missing values from arr_delay - 2.8% of all observations.

```{r}
sum(is.na(flights$arr_delay))
mean(is.na(flights$arr_delay))
```


## (3b) The two mechanisms
There are a total of 9430 missing arrival times, and 8255 missing departure times.

- Mechanism 1: if the flight has a missing arrival and departure, the flight was cancelled.
- Mechanism 2: if the flight has a missing arrival, the flight was rerouted.

```{r}
# total missing values for arrivals and departures
c(miss_arr = sum(is.na(flights$arr_delay)),
  miss_dep = sum(is.na(flights$dep_delay)))

table("miss_arr" = is.na(flights$arr_delay),
      "miss_dep" = is.na(flights$dep_delay))
# Missing none is a completed flight
# Missing both is a cancelled flight
# Missing arrival only is a reroute in the flight
# Missing departure only isn't possible

```


## (3c) What `na.rm = TRUE` actually describes
The average arrival delay of planes that arrived at the planned destination is ~6.89 minutes.
```{r}
mean(flights$arr_delay, na.rm = TRUE)
```


## (3d) Is the deletion carrier-neutral?

- Q) Does na.rm = TRUE treat all carriers equally?
- A) No, na.rm = TRUE does not treat all carriers equally. The function, na.rm = TRUE, removes missing values before the mean is calculated, only those with a recorded arrival time will contribute to the mean. For example, UA is missing 9.5% of arrival times compared to UA who is only missing 3.6% of arrival times. This discrepancy in missing-ness will contribute to the the mean arrival delays for each.
```{r}
flights %>%
  dplyr::group_by(carrier) %>%
  dplyr::summarise(
    n = n(),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    per_na_arr_time = mean(is.na(arr_time) * 100)
  ) %>%
  dplyr::arrange(desc(per_na_arr_time))
```


## (3e) What you would do about it





# Part 4 · A ranking that is not what it looks like (18 points)

## (4a) Carriers by mean arrival delay
```{r}
flights %>%
  dplyr::group_by(carrier) %>%
  dplyr::summarise(
    mean_arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  dplyr::arrange(desc(mean_arr_delay))
```

## (4b) What should stop you in the top five
Of the flights that operated out of New York airports, JFK and LGA, the carrier F9 had the largest mean arrival delay but is missing the fewest arrival times. Compare this to OO which had the smallest mean arrival dealy but is missing the most arrival times. At face value this is showing that OO had more reroutes despite a more timely arrival average, and F9 arrived at the scheduled destination but arrived later on average. Again, this shows how missing values can bias statistics and a subsequent interpretation of the data.

```{r}

table(flights$carrier, flights$origin)

flights %>%
  dplyr::filter(origin %in% c("JFK", "LGA")) %>%
  dplyr::group_by(carrier) %>%
  dplyr::filter(carrier %in% c("F9", "FL", "EV", "YV", "OO")) %>%
  dplyr::summarise(
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    per_na_arr_time = mean(is.na(arr_time) * 100)
  ) %>%
  dplyr::arrange(desc(mean_arr_delay))



```


## (4c) Which panel this is
I would say that panels C and D from "What correlation cannot see" illustrate what is happening with missing-ness in the flight data. Rather than showing a complete picture, select data points (when unaccounted for in an analysis) skew the statistics. Missing data allow existing data to push the statistics in a direction that is not representative of reality.


## (4d) A rule, applied
```{r}
set.seed(27)
# 2727

min_n <- flights %>%
  dplyr::count(carrier) %>%
  dplyr::pull(n) %>%
  min()

rand_flights <- flights %>%
  dplyr::group_by(carrier) %>%
  dplyr::slice_sample(n = min_n) %>%
  dplyr::ungroup()


rand_flights %>%
  dplyr::group_by(carrier) %>%
  dplyr::summarise(
    n = n(),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    per_na_arr_time = mean(is.na(arr_time) * 100)
  ) %>%
  dplyr::arrange(desc(per_na_arr_time))


flights %>%
  dplyr::group_by(carrier) %>%
  dplyr::summarise(
    n = n(),
    mean_arr_delay = mean(arr_delay, na.rm = TRUE),
    per_na_arr_time = mean(is.na(arr_time) * 100)
  ) %>%
  dplyr::arrange(desc(per_na_arr_time))



```




# Part 5 · A plot that misleads, and its honest twin (13 points)

## (5a) The misleading plot

## (5b) The honest plot

## (5c) What changed, and which change did the work

# Part 6 · Who is missing (15 points)

*One paragraph, at most 150 words. Three mechanisms, each with a direction.*

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->