---
title: "Homework 3"
subtitle: "DATS 6101 · Fall 2026"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
data(flights)
```

<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d           rows where both dep_delay and arr_delay are present (327,346 rows)
  late        arr_delay > 15
  a day       the calendar date from year, month, day
  population  d$arr_delay, in Parts 3 to 6; its mean is mu, its sd is sigma
  seeds       set.seed(6101) on the line before each simulation

Part 1 uses all 336,776 rows instead of d. It says so.

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · A payout table (12 points)

## (a) The payout distribution, and its expectation two ways

## (b) Where the expectation sits

## (c) A fair fee, and a 900-flight day

# Part 2 · Sums are free, variances are not (16 points)

## (a) Daily late counts by airport, and the total

## (b) The covariance terms

## (c) For the planner

# Part 3 · A population you can see (18 points)

## (a) mu and sigma

## (b) The sampling distribution at four sample sizes

## (c) One sample of 400: the SD sentence and the SE sentence

## (d) How many flights

# Part 4 · When n is not n (24 points)

## (a) Sixteen days, 2,000 times

## (b) What the formula says, and random flights

## (c) The effective sample size

## (d) The days as the units

## (e) Which step failed

# Part 5 · A rare rate (12 points)

## (a) Two standard errors, and each against its p

## (b) Flights for a 10% relative error

## (c) If the flights came from three days

# Part 6 · The dashboard sentence (8 points)

*One paragraph, at most 150 words.*

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
