---
title: "Homework 3"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
library(tidyverse)
data(flights)

d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

```

<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d           rows where both dep_delay and arr_delay are present (327,346 rows)
  late        arr_delay > 15
  a day       the calendar date from year, month, day
  population  d$arr_delay, in Parts 3 to 6; its mean is mu, its sd is sigma
  seeds       set.seed(6101) on the line before each simulation

Part 1 uses all 336,776 rows instead of d. It says so.

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · A payout table (12 points)
## (1a) The payout distribution, and its expectation two ways
```{r}
# Build payout for every flight
flights <- flights %>%
  mutate(
    payout = case_when(
      is.na(dep_delay) ~ 400,
      !is.na(dep_delay) & is.na(arr_delay) ~ 200,
      arr_delay > 180 ~ 200,
      arr_delay > 60 & arr_delay <= 180 ~ 50,
      TRUE ~ 0
    )
  )

# flights$payout # check column values


# Report the probability of each payout value
pCanceled <- mean(is.na(flights$dep_delay))
pCanceled

pDiverted <- mean(!is.na(flights$dep_delay) & is.na(flights$arr_delay))
pDiverted

p180Late <- sum(flights$arr_delay > 180, na.rm = TRUE) / nrow(flights)
p180Late

pLate <- sum(60 < flights$arr_delay & flights$arr_delay <= 180, na.rm = TRUE) / nrow(flights)
pLate


# Compute E[X] two ways - outputs must agree:

# 1) ... from the table

cashCanceled <- 400
cashDiverted <- 200
cash180Late <- 200
cashLate <- 50
cashOther <- 0 # cancels out

EPayout_theoretical <- (cashCanceled * pCanceled) + (cashDiverted * pDiverted) + (cash180Late * p180Late) + (cashLate * pLate)
EPayout_theoretical

# 2)  ... using mean() of the payout column

EPayout_actual <- mean(flights$payout)
EPayout_actual

isTRUE(all.equal(EPayout_theoretical, EPayout_actual))
EPayout_theoretical == EPayout_actual # Floating point error

```



## (1b) Where the expectation sits
The share of expected payouts from cancellations make up the tail of the distribution, and the expected overall payout sits just past the bulk of data (flights with $0 or $50 payouts) toward the skewed tail of the distribution.

```{r}

pCanceled # share of flights that are canceled

cancelContribtuion <- cashCanceled * pCanceled # payout per flight due to cancellations
cancelShareOfPayout <- cancelContribtuion / EPayout_actual 
cancelShareOfPayout 

```


## (1c) A fair fee, and a 900-flight day
A fair fee is the Expectation of the payout, $16.34, with the expected cost for 900 flights per day being $14,705.95. This Expectation is a linear model, meaning independence/dependence of flights is irrelevant. 

```{r}
E900 <- 900 * EPayout_actual
E900

```



# Part 2 · Sums are free, variances are not (16 points)
## (2a) Daily late counts by airport, and the total

## (2b) The covariance terms

## (2c) For the planner

# Part 3 · A population you can see (18 points)
## (3a) mu and sigma

## (3b) The sampling distribution at four sample sizes

## (3c) One sample of 400: the SD sentence and the SE sentence

## (3d) How many flights

# Part 4 · When n is not n (24 points)
## (4a) Sixteen days, 2,000 times

## (4b) What the formula says, and random flights

## (4c) The effective sample size

## (4d) The days as the units

## (4e) Which step failed

# Part 5 · A rare rate (12 points)
## (5a) Two standard errors, and each against its p

## (5b) Flights for a 10% relative error

## (5c) If the flights came from three days

# Part 6 · The dashboard sentence (8 points)

*One paragraph, at most 150 words.*

# AI disclosure

## Prompt log


