---
title: "Homework 4"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
library(tidyverse)
data(flights)

d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

# A day is a calendar date built from year, month and day.
d <- d %>% dplyr::mutate(date = lubridate::make_date(year, month, day))

population <- d$arr_delay
mu <- mean(population)
sigma <- sd(population)


```

<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d           rows where both dep_delay and arr_delay are present (327,346 rows)
  population  d$arr_delay; its mean is mu, its sd is sigma
  a day       the calendar date from year, month, day
  interval    mean(x) + c(-1, 1) * qt(0.975, n - 1) * sd(x) / sqrt(n)
  coverage    the share of intervals that contain mu -- counted, not assumed
  seeds       set.seed(6101) on the line before each simulation
  reps        2,000 repetitions unless a part says otherwise

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · The theorem is about the mean (12 points)

## (1a) How skewed is arr_delay
Skewness of d$arr_delay: `r skew`.

```{r}
skew <- mean((population - mu)^3) / sigma^3


```


## (1b) The sampling distribution at n = 5, 30, 200
```{r}

```


## (1c) What became Normal, and what did not
```{r}

```


# Part 2 · Counting the coverage (20 points)

## (2a) Coverage at three sample sizes

## (2b) The misses, and which side they are on

## (2c) For the colleague with ten flights

# Part 3 · What the t fixes (10 points)

## (3a) The same intervals with 1.96

## (3b) What the t corrects for, and what it cannot

# Part 4 · The interval that is not about the population (20 points)

## (4a) One whole day, two thousand times

## (4b) The same number of random flights

## (4c) Which step failed

## (4d) The fix, measured

# Part 5 · An interval for a rare rate (16 points)

## (5a) p, and np at n = 200

## (5b) The textbook interval

## (5c) prop.test

## (5d) What you would report

# Part 6 · The sentence, and what precision costs (12 points)

## (6a) One sample, one interval

## (6b) The sentence, and two that are wrong

## (6c) The cost of halving the width

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
