---
title: "Homework 4"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---


https://www.smajhi.com/DATS-6101/homework/hw4.html#ai-disclosure-required

https://claude.ai/chat/b170d8f8-6fcf-4a6f-b8c1-06f55734c649

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
library(tidyverse)
data(flights)

d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

# A day is a calendar date built from year, month and day.
d <- d %>% dplyr::mutate(date = lubridate::make_date(year, month, day))

population <- d$arr_delay
mu <- mean(population)
sigma <- sd(population)

```
Confirm observation count for for `d`: `r nrow(d)`

Population defined as `d$arr_delay`

Mu: `r round(mean(population), 2)`

Sigma: `r round(sd(population), 2)`


<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d           rows where both dep_delay and arr_delay are present (327,346 rows)
  population  d$arr_delay; its mean is mu, its sd is sigma
  a day       the calendar date from year, month, day
  interval    mean(x) + c(-1, 1) * qt(0.975, n - 1) * sd(x) / sqrt(n)
  coverage    the share of intervals that contain mu -- counted, not assumed
  seeds       set.seed(6101) on the line before each simulation
  reps        2,000 repetitions unless a part says otherwise

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · The theorem is about the mean (12 points)

## (1a) How skewed is arr_delay
```{r}
skewness <- function(x) mean((x - mean(x))^3) / sd(x)^3 # define skew function

pop_skew <- skewness(population)

```
Skewness of `arr_delay`: `r round(pop_skew, 2)`.


## (1b) The sampling distribution at n = 5, 30, 200
```{r}

# Define sample sizes and replicates
n5 <- 5
n30 <- 30
n200 <- 200
reps <- 5000

# Define sample function for simulation replicability
# n = sample size
# reps = number of replications to perform for simulation
# pop = population to sample from
sim_means <- function(n, reps, pop) {
  set.seed(6101)
  replicate(reps, mean(sample(pop, n, replace = TRUE)))
}

# 5000 means of sample size, n
mean_5 <- sim_means(n5, reps, population) 
mean_30 <- sim_means(n30, reps, population)
mean_200 <- sim_means(n200, reps, population)

# all means of 5000 of sample size, n
all_means_5000 <- tibble(
  n = rep(c(n5, n30, n200), each = reps),
  xbar = c(mean_5, mean_30, mean_200))
head(all_means_5000, 5)

# grand means and associated stats
grand_stats <- tibble(
  n = c(n5, n30, n200),
  means = c(mean(mean_5), mean(mean_30), mean(mean_200)), # grand means at sample size n
  sd = c(sd(mean_5), sd(mean_30), sd(mean_200)), # sd of means at sample size n
  skew = c(skewness(mean_5), skewness(mean_30), skewness(mean_200)) # skew of means at sample size n
)

# 3 histograms
ggplot(all_means_5000, aes(xbar)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ n, nrow = 1, scales = "free_x", labeller = label_both) +
  labs(title = "Histograms of 5000 Simulated Sampled Means at: \n n = 5, 30, 200", 
       x = "Value of sampled means (minutes)", 
       y = "Count of means")

```


## (1c) What became Normal, and what did not
The skewness for each sample size (5, 30, 200) are: `r round(grand_stats[1, 4], 2)`, `r round(grand_stats[2, 4], 2)`, and `r round(grand_stats[3, 4], 2)`, respectively. Sample size, n = 5, is the most skewed, resembling the population from which it was sampled (`r round(pop_skew, 2)`). As sample size increases, the distribution converges on normality, where n = 200 is closest to normality.

# Part 2 · Counting the coverage (20 points)

## (2a) Coverage at three sample sizes
```{r}
n10 <- 10
reps2 <- 2000


sim_t_stats <- function(n, reps, pop) {
  set.seed(6101)
  out <- replicate(reps, { # specify replicates
    x <- sample(pop, n, replace = TRUE) # draw samples of size n
    c(xbar = mean(x), s = sd(x)) # calculate mean and sd of each sample of size n
  })
  tibble(xbar = out["xbar", ], s = out["s", ]) # output as tibble
}

sim_2k_means10  <- sim_t_stats(n10,  reps2, population)
sim_2k_means30  <- sim_t_stats(n30,  reps2, population)
sim_2k_means200 <- sim_t_stats(n200, reps2, population)


# Function to get confidence intervals for each sample and if sample CI includes:
# stats is the tibble from sim_t_stats with mean and sd
# n is sample of size n
# mu is population mean
t_coverage <- function(stats, n, mu) {
  half <- qt(0.975, n - 1) * stats$s / sqrt(n) # calculate half of confidence interval for each sample of size n
  mean(stats$xbar - half <= mu & mu <= stats$xbar + half) # proportions of confidence intervals that contain mu
}

# Coverage for each sample of size n
cov_10  <- t_coverage(sim_2k_means10,  n10,  mu)
cov_30  <- t_coverage(sim_2k_means30,  n30,  mu)
cov_200 <- t_coverage(sim_2k_means200, n200, mu)
cov_10
cov_30
cov_200

```
For each sample size (n = 10, 30, 200) at 2000 replications, the respective coverages are:

n = 10: `r cov_10`

n = 30: `r cov_30`

n = 200: `r cov_200`


## (2b) The misses, and which side they are on
```{r}
# Function to get proportion of sample CIs where CI is below MU
# stats is the tibble from sim_t_stats with mean and sd
# n is sample of size n
# mu is population mean
t_miss_summary <- function(stats, n, mu) {
  margin <- qt(0.975, n - 1) * stats$s / sqrt(n)  # half-width of each interval
  lower  <- stats$xbar - margin
  upper  <- stats$xbar + margin

  covered    <- lower <= mu & mu <= upper # is mu included in CI
  miss_below <- upper < mu      # whole interval left of mu
  miss_above <- lower > mu      # whole interval right of mu

  tibble(
    n           = n,
    coverage    = mean(covered), # proportion of samples that include mu
    n_miss      = sum(!covered), # proportion of samples that exclude mu
    share_below = sum(miss_below) / sum(!covered) # proportion of samples where CI is below MU
  )
}

miss_tbl <- bind_rows(
  t_miss_summary(sim_2k_means10,  n10,  mu),
  t_miss_summary(sim_2k_means30,  n30,  mu),
  t_miss_summary(sim_2k_means200, n200, mu)
)

miss_tbl

```
None of the given sample sizes achieved 95% coverage. All fell below, with n = 200 coming closest at `r cov_200`.
For each sample CI, the majority of CIs that did not include mu, were below mu, because the the extreme values from a right-skewed distribution pull mu to the right. 

## (2c) For the colleague with ten flights
```{r}
# Function to get confidence intervals for each sample and if sample CI includes:
# stats is the tibble from sim_t_stats with mean and sd
# n is sample of size n
# mu is population mean
t_99_coverage <- function(stats, n, mu) {
  half <- qt(0.995, n - 1) * stats$s / sqrt(n) # calculate half of confidence interval for each sample of size n
  mean(stats$xbar - half <= mu & mu <= stats$xbar + half) # proportions of confidence intervals that contain mu
}

# Coverage for each sample of size n
cov_10_99  <- t_99_coverage(sim_2k_means10,  n10,  mu)
cov_30_99  <- t_99_coverage(sim_2k_means30,  n30,  mu)
cov_200_99 <- t_99_coverage(sim_2k_means200, n200, mu)
cov_10_99
cov_30_99
cov_200_99


```

Increasing the confidence level will widen the confidence interval, but precision is the cost in estimating mu from widening the interval. For example, at 95% confidence the coverage is as stated previously: `r round(cov_10,2)`, `r round(cov_30, 2)`, `r round(cov_200, 2)` (n = 10, 30 200). When confidence is increased, the proportion of intervals that contain mu increases: `r cov_10_99`, `r cov_30_99`, `r cov_200_99` (n = 10, 30, 200).


# Part 3 · What the t fixes (10 points)

## (3a) The same intervals with 1.96
```{r}

# Function to get confidence intervals using qnorm() for each sample and if sample CI:
# stats is the tibble from sim_t_stats with mean and sd
# n is sample of size n
# mu is population mean
norm_coverage <- function(stats, n, mu) {
  half <- qnorm(0.975) * stats$s / sqrt(n) # calculate half of confidence interval for each sample of size n
  mean(stats$xbar - half <= mu & mu <= stats$xbar + half) # proportions of confidence intervals that contain mu
}

cov_10
cov_30
cov_200


norm_cov_10 <- norm_coverage(sim_2k_means10,  n10,  mu)
norm_cov_10
norm_cov_30 <- norm_coverage(sim_2k_means30,  n30,  mu)
norm_cov_30
norm_cov_200 <- norm_coverage(sim_2k_means200,  n200,  mu)
norm_cov_200
```

## (3b) What the t corrects for, and what it cannot
The t-distribution corrects for a narrower confidence interval that would be generated by using the normal distribution. Howeever, the t distribution can't correct for data that are skewed (as in this example), if anything the assumption of normality is stricter with a t-distribution but is robust against deviation from normality as n grows which is also why CIs generated for n = 200 agree more at this sample size. 

# Part 4 · The interval that is not about the population (20 points)

## (4a) One whole day, two thousand times
```{r}

# one row per calendar date: flights that day, mean and sd of arr_delay
day_stats <- d %>%
  group_by(date) %>%
  summarise(n = n(), xbar = mean(arr_delay), s = sd(arr_delay))

set.seed(6101)

# reps2 = 2000 from above
sample_days <- day_stats[sample(nrow(day_stats), reps2, replace = TRUE), ] 

ci_days_half    <- qt(0.975, sample_days$n - 1) * sample_days$s / sqrt(sample_days$n)  # half of CI
ci_days_lower   <- sample_days$xbar - ci_days_half # lower CI bound
ci_days_upper   <- sample_days$xbar + ci_days_half # upper CI bound
covered_days <- ci_days_lower <= mu & mu <= ci_days_upper # is mu in interval

avg_flights <- mean(sample_days$n) # average flights per interval
avg_flights
cov_days     <- mean(covered_days) # share of intervals that contain mu
cov_days
avg_width   <- mean(ci_days_upper - ci_days_lower)   # average width of intervals
avg_width

```
Average number of flights per interval: `r round(avg_flights, 2)`

Proportion of CIs that include mu: `r round(cov_days, 2)`

Average width of intervals: `r round(avg_width, 2)`


## (4b) The same number of random flights
```{r}

set.seed(6101)

# one random sample from the whole population for each day-interval's sample size
pop_stats <- sapply(sample_days$n, function(k) { # same number of flights
  x <- sample(population, k, replace = TRUE) # sample from entire population
  c(xbar = mean(x), s = sd(x)) # xbar and sd for each sample
})   # 2 x 2000 matrix: row "xbar" and row "s"

ci_pop_half  <- qt(0.975, sample_days$n - 1) * pop_stats["s", ] / sqrt(sample_days$n) # half width of CI
ci_pop_lower <- pop_stats["xbar", ] - ci_pop_half
ci_pop_upper <- pop_stats["xbar", ] + ci_pop_half
covered_pop  <- ci_pop_lower <= mu & mu <= ci_pop_upper

cov_pop       <- mean(covered_pop)                 # share of intervals that contain mu
avg_width_pop <- mean(ci_pop_upper - ci_pop_lower)


# compare whole day sampling to full population sampling
compare_days <- tibble(
  sampling     = c("One whole day", "Random flights, whole file"),
  avg_flights  = c(mean(sample_days$n), mean(sample_days$n)),
  coverage     = c(cov_days, cov_pop),
  avg_width    = c(avg_width, avg_width_pop)
)
knitr::kable(compare_days, digits = 2)

se_days <- mean(sample_days$s / sqrt(sample_days$n))  # average SE, day intervals
se_pop  <- mean(pop_stats["s", ] / sqrt(sample_days$n))  # average SE, random samples


```


## (4c) Which step failed
Sampling a whole day lacks independence among flights (step 2 fails in the derivation), so the variance within a day shrinks and the standard error is too small (`r round(se_days, 2)`. The CI is too narrow and centered on a day-specific mean that is often far from mu (the interval is for the day not for mu), so it rarely captures mu (`r round(cov_days, 2)`). Widening in the intervals won't help as a result since the days interval is so far from mu to begin with. Truly independent sampling captured mu in `r round(cov_pop, 2)` of intervals, close to the nominal 95%.

## (4d) The fix, measured
```{r}
# the 365 daily means (one number per calendar date)
daily_means <- day_stats$xbar
head(daily_means, 10)

n16   <- 16


# draw 16 days at random, build the t interval from their 16 means
# (replace = FALSE: 16 distinct days from the 365)
# n = sample size
# reps = replications of simulation
# dm = daily means

sim_day_ci <- function(n, reps, dm) {
  set.seed(6101)
  out <- replicate(reps, {
    x <- sample(dm, n)
    c(xbar = mean(x), s = sd(x))
  })
  tibble(xbar = out["xbar", ], s = out["s", ])
}

day16_stats <- sim_day_ci(n16, reps2, daily_means)

ci16_half  <- qt(0.975, n16 - 1) * day16_stats$s / sqrt(n16)
ci16_lower <- day16_stats$xbar - ci16_half
ci16_upper <- day16_stats$xbar + ci16_half
covered_16 <- ci16_lower <= mu & mu <= ci16_upper

nrow(day16_stats)

cov_16        <- mean(covered_16)   # share of intervals that contain mu
cov_16
avg_half_16   <- mean(ci16_half)    # average half-width
avg_half_16

```
The coverage from randomly sampling 16 days has substantially improved (`r cov_16`, closer to the desired 95% confidence) compared to sampling whole days (`r cov_days`).


# Part 5 · An interval for a rare rate (16 points)

## (5a) p, and np at n = 200
```{r}

# proportion of flights arrive more than 3 hours late
p <- mean(d$arr_delay > 180)
p

n200 <- 200

np200 <- n200 * p
np200

```


## (5b) The textbook interval
Draw 2,000 samples of 200 flights. For each, build the textbook proportion interval, p^±1.96p^(1−p^)/np^​±1.96p^​(1−p^​)/n, and report: its coverage, the share of samples containing no such flight at all, and the share whose lower bound is negative.

```{r}


# returns the sample share of flights arriving more than 180 minutes late, one per sample
# n = sample size
# reps = number of replications
# pop = population to sample from

sim_props <- function(n, reps, pop) {
  set.seed(6101)
  replicate(reps, mean(sample(pop, n, replace = TRUE) > 180))
}

phat200 <- sim_props(n200, reps2, population)


se_phat    <- sqrt(phat200 * (1 - phat200) / n200) # SE for proportion
ci_p_lower <- phat200 - 1.96 * se_phat
ci_p_upper <- phat200 + 1.96 * se_phat

covered_p  <- ci_p_lower <= p & p <= ci_p_upper # is p in interval

cov_p      <- mean(covered_p)           # share of intervals containing p
cov_p
share_none <- mean(phat200 == 0)        # share of samples with no late flight at all
share_none
share_neg  <- mean(ci_p_lower < 0)      # share of intervals with a negative lower bound
share_neg




```


## (5c) prop.test
```{r}

# count of late flights in each sample, recovered from the sample share
k200 <- round(phat200 * n200)

# prop.test interval for every sample
ci_pt <- sapply(k200, function(k) prop.test(k, n200)$conf.int)

ci_pt_lower <- ci_pt[1, ]
ci_pt_upper <- ci_pt[2, ]

covered_pt <- ci_pt_lower <= p & p <= ci_pt_upper
cov_pt     <- mean(covered_pt)   # share of intervals that contain p
cov_pt

cov_ptest <- prop.test(0, 200)$conf.int
cov_ptest
cov_ptest[1]
cov_ptest[2]

```

## (5d) What you would report
Flights that are more than 3 hours late are very rare making the proportion essentially 0, but the interval goes into negative numbers for samples with only a couple of late flights, which doesn't make any sense given the context. Overall the entirety of the interval is a single point at 0. The function prop.test() is more appropriate for events that are rare, like this one. So when reporting to a manager for the rate, I would use prop.test() to give `r cov_pt` (`r cov_ptest[1]`, `r cov_ptest[2]`)

# Part 6 · The sentence, and what precision costs (12 points)

## (6a) One sample, one interval

## (6b) The sentence, and two that are wrong

## (6c) The cost of halving the width

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
