---
title: "Homework 4"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
library(tidyverse)
data(flights)

d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

# A day is a calendar date built from year, month and day.
d <- d %>% dplyr::mutate(date = lubridate::make_date(year, month, day))

population <- d$arr_delay
mu <- mean(population)
sigma <- sd(population)

```
Confirm observation count for for `d`: `r nrow(d)`

Population defined as `d$arr_delay`

Mu: `r round(mean(population), 2)`

Sigma: `r round(sd(population), 2)`


<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d           rows where both dep_delay and arr_delay are present (327,346 rows)
  population  d$arr_delay; its mean is mu, its sd is sigma
  a day       the calendar date from year, month, day
  interval    mean(x) + c(-1, 1) * qt(0.975, n - 1) * sd(x) / sqrt(n)
  coverage    the share of intervals that contain mu -- counted, not assumed
  seeds       set.seed(6101) on the line before each simulation
  reps        2,000 repetitions unless a part says otherwise

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · The theorem is about the mean (12 points)

## (1a) How skewed is arr_delay
```{r}
skewness <- function(x) mean((x - mean(x))^3) / sd(x)^3 # define skew function

pop_skew <- skewness(population)

```
Skewness of `arr_delay`: `r round(pop_skew, 2)`.


## (1b) The sampling distribution at n = 5, 30, 200
```{r}

# Define sample sizes and replicates
n5 <- 5
n30 <- 30
n200 <- 200
reps <- 5000

# Define sample function for simulation replicability
# n = sample size
# reps = number of replications to perform for simulation
# pop = population to sample from
sim_means <- function(n, reps, pop) {
  set.seed(6101)
  replicate(reps, mean(sample(pop, n, replace = TRUE)))
}

# 5000 means of sample size, n
mean_5 <- sim_means(n5, reps, population) 
mean_30 <- sim_means(n30, reps, population)
mean_200 <- sim_means(n200, reps, population)

# all means of 5000 of sample size, n
all_means_5000 <- tibble(
  n = rep(c(n5, n30, n200), each = reps),
  xbar = c(mean_5, mean_30, mean_200))
head(all_means_5000, 5)

# grand means and associated stats
grand_stats <- tibble(
  n = c(n5, n30, n200),
  means = c(mean(mean_5), mean(mean_30), mean(mean_200)), # grand means at sample size n
  sd = c(sd(mean_5), sd(mean_30), sd(mean_200)), # sd of means at sample size n
  skew = c(skewness(mean_5), skewness(mean_30), skewness(mean_200)) # skew of means at sample size n
)

# 3 histograms
ggplot(all_means_5000, aes(xbar)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ n, nrow = 1, scales = "free_x", labeller = label_both) +
  labs(title = "Histograms of 5000 Simulated Sampled Means at: \n n = 5, 30, 200", 
       x = "Value of sampled means (minutes)", 
       y = "Count of means")

```


## (1c) What became Normal, and what did not
The skewness for each sample size (5, 30, 200) are: `r round(grand_stats[1, 4], 2)`, `r round(grand_stats[2, 4], 2)`, and `r round(grand_stats[3, 4], 2)`, respectively. Sample size, n = 5, is the most skewed, resembling the population from which it was sampled (`r round(pop_skew, 2)`). As sample size increases, the distribution converges on normality, where n = 200 is closest to normality.

# Part 2 · Counting the coverage (20 points)

## (2a) Coverage at three sample sizes
```{r}
n10 <- 10
reps2 <- 2000


sim_t_stats <- function(n, reps, pop) {
  set.seed(6101)
  out <- replicate(reps, { # specify replicates
    x <- sample(pop, n, replace = TRUE) # draw samples of size n
    c(xbar = mean(x), s = sd(x)) # calculate mean and sd of each sample of size n
  })
  tibble(xbar = out["xbar", ], s = out["s", ]) # output as tibble
}

sim_2k_means10  <- sim_t_stats(n10,  reps2, population)
sim_2k_means30  <- sim_t_stats(n30,  reps2, population)
sim_2k_means200 <- sim_t_stats(n200, reps2, population)


# Function to get confidence intervals for each sample and if sample CI includes:
# stats is the tibble from sim_t_stats with mean and sd
# n is sample of size n
# mu is population mean
t_coverage <- function(stats, n, mu) {
  half <- qt(0.975, n - 1) * stats$s / sqrt(n) # calculate half of confidence interval for each sample of size n
  mean(stats$xbar - half <= mu & mu <= stats$xbar + half) # proportions of confidence intervals that contain mu
}

# Coverage for each sample of size n
cov_10  <- t_coverage(sim_2k_means10,  n10,  mu)
cov_30  <- t_coverage(sim_2k_means30,  n30,  mu)
cov_200 <- t_coverage(sim_2k_means200, n200, mu)


```
For each sample size (n = 10, 30, 200) at 2000 replications, the respective coverages are:
n = 10: `r cov_10`

n = 30: `r cov_30`

n = 200:`r cov_200`


## (2b) The misses, and which side they are on

## (2c) For the colleague with ten flights

# Part 3 · What the t fixes (10 points)

## (3a) The same intervals with 1.96

## (3b) What the t corrects for, and what it cannot

# Part 4 · The interval that is not about the population (20 points)

## (4a) One whole day, two thousand times

## (4b) The same number of random flights

## (4c) Which step failed

## (4d) The fix, measured

# Part 5 · An interval for a rare rate (16 points)

## (5a) p, and np at n = 200

## (5b) The textbook interval

## (5c) prop.test

## (5d) What you would report

# Part 6 · The sentence, and what precision costs (12 points)

## (6a) One sample, one interval

## (6b) The sentence, and two that are wrong

## (6c) The cost of halving the width

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
