---
title: "Homework 2"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
library(tidyverse)
data(flights)
```

<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d        rows where both dep_delay and arr_delay are present (327,346 rows)
  A        arr_delay > 15
  D        dep_delay > 15
  a day    the calendar date from year, month, day
  morning  dep_time %/% 100 below 12;  evening: 17 or later

Part 4(a) uses all 336,776 rows instead of d. It says so.

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · Events are logical vectors (10 points)

https://www.smajhi.com/DATS-6101/homework/hw2.html

## (1a) P(A), P(D), P(A and D)

```{r}
d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

A <- d$arr_delay > 15 # flights that arrived >15 min late
head(A, 50) # check vector

D <- d$dep_delay > 15  # flights that departed >15 min late
head(D, 50) # check vector

```



## (1b) P(A or D), two ways

```{r}
# P(AuD), probability that flight is late by arrival OR by departure

pA <- mean(d$arr_delay > 15, na.rm = TRUE) # P(late arrival)
pD <- mean(d$dep_delay > 15, na.rm = TRUE) # P(late departure)

# By operator '|'
pAuD_opp <- mean(A | D, na.rm = TRUE)

# By additive rule
pAnD_opp <- mean(A & D, na.rm = TRUE) # calculate p(AnD)
pAuD_add <- pA + pD - pAnD_opp # calculate p(AuD) by addititve rule

isTRUE(all.equal(pAuD_opp, pAuD_add)) # vectors are the same mathematically

pAuD_opp == pAuD_add # vectors differ by float-point
print(pAuD_opp, digits = 20)
print(pAuD_add, digits = 20)

```



## (1c) Neither
The complement of p(AuD) gets us to p(Ac n Dc).
1 - p(AuD) = p(Ac n Dc)

```{r}
# P(flight has on time departure and arrival)
# Complement of p(AuD)
cAuD <- 1 - pAuD_add
cAuD

```




# Part 2 · Independence, checked rather than assumed (18 points)

## (2a) The product, the ratio, and P(A | D)
When a flight is pushed back (late departure), the flight will most likley arrive late at the destination.

```{r}

# Check A and D independence
pApD <- pA * pD 
isTRUE(all.equal(pApD, pAuD_opp))
pAuD_opp
pApD

pAuD_opp / pApD # p(AuD) is ~5.5 times larger than p(A)p(D)

# Compute p(A|D)
# p(A|D) = p(AnD) / p(D)

pAgD <- pAnD_opp / pD
pAgD
```


## (2b) Late, by origin

*Numbers first, then whether the difference matters.*

p(A) = 0.237
p(A | EWR) = 0.256
p(A | JFK) = 0.23
p(A | LGA) = 0.224

p(A | EWR) / p(A) = 1.079
p(A | JFK) / p(A) = 0.968
p(A | LGA) / p(A) = 0.943

The comparison of a late arrival given a specific airport or origin to overall late arrival for the NYC region is not meaningfully different.
Travelers would not gain an advantage in timeliness from one airport to another.

```{r}

# Compute p(Origin)
origin_ewr <- d$origin == "EWR"
head(origin_ewr, 50)
pEWR <- mean(origin_ewr)
pEWR

origin_jfk <- d$origin == "JFK"
head(origin_jfk, 50)
pJFK <- mean(origin_jfk)
pJFK

origin_lga <- d$origin == "LGA"
head(origin_lga, 50)
pLGA <- mean(origin_lga)
pLGA

# Vectors should not be the same
isTRUE(all.equal(origin_ewr, origin_jfk))
isTRUE(all.equal(origin_ewr, origin_lga))
isTRUE(all.equal(origin_jfk, origin_lga))

# Compute p(A | Origin)
# Subset proportion of late arrivals by origin
pAgEWR <- mean(A[origin_ewr])
pAgJFK <- mean(A[origin_jfk])
pAgLGA <- mean(A[origin_lga])


pA # 0.237
pAgEWR # 0.256
pAgJFK # 0.23
pAgLGA # 0.224

pAgEWR / pA # 1.079
pAgJFK / pA # 0.968
pAgLGA / pA # 0.943
# Late arrivals to not meaningfully differ by airport relative to overall late arrivals

```

## (2c) Two flights, same day
Correlation: 0.6321228
Avg morning prop: 0.1459389
Avg evening prop: 0.3366757
Product of avgs: 0.04913409

```{r}

min(d$sched_dep_time)
max(d$sched_dep_time)

# 2c-a) Compute proportion of late morning, and late evening flights for each day
# Create variables for morning/evening, late: T/F
# Morning flight: 500 - 1159
# Evening flight: 1700 - 2359
d <- d %>%
  mutate(
    time_of_day = case_when(
      sched_dep_time >= 500 & sched_dep_time <= 1159 ~ "morning",
      sched_dep_time >= 1700 & sched_dep_time <= 2359 ~ "evening",
      TRUE ~ "afternoon"),
    late = arr_delay > 15)

table(d$time_of_day)
table(d$late)

late_by_day <- d %>%
  filter(time_of_day %in% c("morning", "evening")) %>%
  group_by(day, time_of_day) %>%
  summarise(prop_late = mean(late), n = n(), .groups = "drop")

late_by_day




# 2c-b) Correlation between morning and evening flights across days
# transpose to wide to get morning/evening columns
late_wide <- late_by_day %>%
  select(day, time_of_day, prop_late) %>%
  pivot_wider(names_from = time_of_day, values_from = prop_late)

plot(late_wide$morning, late_wide$evening) # plot
cor(late_wide$morning, late_wide$evening) # correlation coefficient


# 2c-c) Report the average morning proportion, average evening proportion, and their product
avg_even_late <- mean(late_by_day$prop_late[late_by_day$time_of_day == "evening"])
median(late_by_day$prop_late[late_by_day$time_of_day == "evening"])
sd(late_by_day$prop_late[late_by_day$time_of_day == "evening"])
IQR(late_by_day$prop_late[late_by_day$time_of_day == "evening"])

avg_morn_late <- mean(late_by_day$prop_late[late_by_day$time_of_day == "morning"])
median(late_by_day$prop_late[late_by_day$time_of_day == "morning"])
sd(late_by_day$prop_late[late_by_day$time_of_day == "morning"])
IQR(late_by_day$prop_late[late_by_day$time_of_day == "morning"])


avg_morn_late * avg_even_late
# 2c-d) Estimate p(Morning Flight late & Evening Flight late on same day) as average over days
average_of_prod <- mean(late_wide$morning * late_wide$evening) # p(late morning & evening | any given day)
prod_of_averages <- avg_morn_late * avg_even_late # p(late morning AND late evening) assuming independence
average_of_prod
prod_of_averages

average_of_prod - prod_of_averages
# small difference in probabilities = not mathematically independent 
# demonstrates independence of morning vs evening lateness in principle


```


# Part 3 · A rule that works, and the same rule that does not (22 points)

## (3a) Arriving more than 15 minutes late
```{r}

flag <- d$dep_delay > 30
act_late <- d$arr_delay > 15

pF <- mean(flag)
pL <- pA

table(Flagged = flag, 
      Late = act_late)

# Sensitivity: True positive, probability that the flight was flagged given that it was actually late
#   p(F | L)
pFgL <- mean(flag[act_late])
pFgL

# Specificity: True negative, probability that the flight wasn't flagged given that it was on time
#   p(Fc | Lc)
pFcgLc <- mean(!flag[!act_late])
pFcgLc

# Precision: Probability that the flight was actually late, given that it was flagged
#   p(L | F)
pLgF <- mean(act_late[flag])
pLgF


```


## (3b) Arriving more than three hours late
```{r}

# New flag rule > 180
F180 <- d$dep_delay > 180

pF180 <- mean(F180)

# Sensitivity: True positive, probability that the flight was flagged given that it was actually late
#   p(F | L)

pF180gL <- mean(F180[act_late])
pF180gL

# Specificity: True negative, probability that the flight wasn't flagged given that it was on time
#   p(Fc | Lc)
pF180cgLc <- mean(!F180[!act_late])
pF180cgLc

# Precision: Probability that the flight was actually late, given that it was flagged
#   p(L | F)
pLgF180 <- mean(act_late[F180])
pLgF180

```



## (3c) The same precision, by Bayes' rule
```{r}

```


## (3d) For the manager

# Part 4 · Which distribution, and why not (22 points)

## (4a) Cancellations and scheduled departures

## (4b) Late flights per day, JFK to LAX

## (4c) The 68-95-99.7 rule, twice

# Part 5 · < or ≤ (8 points)

## (5a) The daily late count

## (5b) air_time

# Part 6 · The staffing sentence (10 points)

*One paragraph, at most 150 words.*

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
