---
title: "Homework 2"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
library(tidyverse)
data(flights)
```

<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d        rows where both dep_delay and arr_delay are present (327,346 rows)
  A        arr_delay > 15
  D        dep_delay > 15
  a day    the calendar date from year, month, day
  morning  dep_time %/% 100 below 12;  evening: 17 or later

Part 4(a) uses all 336,776 rows instead of d. It says so.

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · Events are logical vectors (10 points)

https://www.smajhi.com/DATS-6101/homework/hw2.html

## (1a) P(A), P(D), P(A and D)

```{r}
d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

A <- d$arr_delay > 15 # flights that arrived >15 min late
head(A, 50) # check vector

D <- d$dep_delay > 15  # flights that departed >15 min late
head(D, 50) # check vector

```



## (1b) P(A or D), two ways

```{r}
# P(AuD), probability that flight is late by arrival OR by departure

pA <- mean(d$arr_delay > 15, na.rm = TRUE) # P(late arrival)
pD <- mean(d$dep_delay > 15, na.rm = TRUE) # P(late departure)

# By operator '|'
pAuD_opp <- mean(A | D, na.rm = TRUE)

# By additive rule
pAnD_opp <- mean(A & D, na.rm = TRUE) # calculate p(AnD)
pAuD_add <- pA + pD - pAnD_opp # calculate p(AuD) by addititve rule

isTRUE(all.equal(pAuD_opp, pAuD_add)) # vectors are the same mathematically

pAuD_opp == pAuD_add # vectors differ by float-point
print(pAuD_opp, digits = 20)
print(pAuD_add, digits = 20)

```



## (1c) Neither
The complement of p(AuD) gets us to p(Ac n Dc).
1 - p(AuD) = p(Ac n Dc)

```{r}
# P(flight has on time departure and arrival)
# Complement of p(AuD)
cAuD <- 1 - pAuD_add
cAuD

```




# Part 2 · Independence, checked rather than assumed (18 points)

## (2a) The product, the ratio, and P(A | D)
When a flight is pushed back (late departure), the flight will most likley arrive late at the destination.

```{r}

# Check A and D independence
pApD <- pA * pD 
isTRUE(all.equal(pApD, pAuD_opp))
pAuD_opp
pApD

pAuD_opp / pApD # p(AuD) is ~5.5 times larger than p(A)p(D)

# Compute p(A|D)
# p(A|D) = p(AnD) / p(D)

pAgD <- pAnD_opp / pD
pAgD
```


## (2b) Late, by origin

*Numbers first, then whether the difference matters.*

p(A) = 0.237
p(A | EWR) = 0.256
p(A | JFK) = 0.23
p(A | LGA) = 0.224

p(A | EWR) / p(A) = 1.079
p(A | JFK) / p(A) = 0.968
p(A | LGA) / p(A) = 0.943

The comparison of a late arrival given a specific airport or origin to overall late arrival for the NYC region is not meaningfully different.
Travelers would not gain an advantage in timeliness from one airport to another.

```{r}

# Compute p(Origin)
origin_ewr <- d$origin == "EWR"
head(origin_ewr, 50)
pEWR <- mean(origin_ewr)
pEWR

origin_jfk <- d$origin == "JFK"
head(origin_jfk, 50)
pJFK <- mean(origin_jfk)
pJFK

origin_lga <- d$origin == "LGA"
head(origin_lga, 50)
pLGA <- mean(origin_lga)
pLGA

# Vectors should not be the same
isTRUE(all.equal(origin_ewr, origin_jfk))
isTRUE(all.equal(origin_ewr, origin_lga))
isTRUE(all.equal(origin_jfk, origin_lga))

# Compute p(A | Origin)
# Subset proportion of late arrivals by origin
pAgEWR <- mean(A[origin_ewr])
pAgJFK <- mean(A[origin_jfk])
pAgLGA <- mean(A[origin_lga])


pA # 0.237
pAgEWR # 0.256
pAgJFK # 0.23
pAgLGA # 0.224

pAgEWR / pA # 1.079
pAgJFK / pA # 0.968
pAgLGA / pA # 0.943
# Late arrivals to not meaningfully differ by airport relative to overall late arrivals

```

## (2c) Two flights, same day
Correlation: 0.6321228
Avg morning prop: 0.1459389
Avg evening prop: 0.3366757
Product of avgs: 0.04913409

```{r}

min(d$sched_dep_time)
max(d$sched_dep_time)

# 2c-a) Compute proportion of late morning, and late evening flights for each day
# Create variables for morning/evening, late: T/F
# Morning flight: 500 - 1159
# Evening flight: 1700 - 2359
d <- d %>%
  mutate(
    time_of_day = case_when(
      sched_dep_time >= 500 & sched_dep_time <= 1159 ~ "morning",
      sched_dep_time >= 1700 & sched_dep_time <= 2359 ~ "evening",
      TRUE ~ "afternoon"),
    late = arr_delay > 15)

table(d$time_of_day)
table(d$late)

late_by_day <- d %>%
  filter(time_of_day %in% c("morning", "evening")) %>%
  group_by(day, time_of_day) %>%
  summarise(prop_late = mean(late), n = n(), .groups = "drop")

late_by_day




# 2c-b) Correlation between morning and evening flights across days
# transpose to wide to get morning/evening columns
late_wide <- late_by_day %>%
  select(day, time_of_day, prop_late) %>%
  pivot_wider(names_from = time_of_day, values_from = prop_late)

plot(late_wide$morning, late_wide$evening) # plot
cor(late_wide$morning, late_wide$evening) # correlation coefficient


# 2c-c) Report the average morning proportion, average evening proportion, and their product
avg_even_late <- mean(late_by_day$prop_late[late_by_day$time_of_day == "evening"])
median(late_by_day$prop_late[late_by_day$time_of_day == "evening"])
sd(late_by_day$prop_late[late_by_day$time_of_day == "evening"])
IQR(late_by_day$prop_late[late_by_day$time_of_day == "evening"])

avg_morn_late <- mean(late_by_day$prop_late[late_by_day$time_of_day == "morning"])
median(late_by_day$prop_late[late_by_day$time_of_day == "morning"])
sd(late_by_day$prop_late[late_by_day$time_of_day == "morning"])
IQR(late_by_day$prop_late[late_by_day$time_of_day == "morning"])


avg_morn_late * avg_even_late
# 2c-d) Estimate p(Morning Flight late & Evening Flight late on same day) as average over days
average_of_prod <- mean(late_wide$morning * late_wide$evening) # p(late morning & evening | any given day)
prod_of_averages <- avg_morn_late * avg_even_late # p(late morning AND late evening) assuming independence
average_of_prod
prod_of_averages

average_of_prod - prod_of_averages
# small difference in probabilities = not mathematically independent 
# demonstrates independence of morning vs evening lateness in principle


```


# Part 3 · A rule that works, and the same rule that does not (22 points)

## (3a) Arriving more than 15 minutes late
```{r}

flag <- d$dep_delay > 30
act_late <- d$arr_delay > 15

pF <- mean(flag)
pL <- pA

table(Flagged = flag, 
      Late = act_late)

# Sensitivity: True positive, probability that the flight was flagged given that it was actually late
#   p(F | L)
pFgL <- mean(flag[act_late])
pFgL

# Specificity: True negative, probability that the flight wasn't flagged given that it was on time
#   p(Fc | Lc)
pFcgLc <- mean(!flag[!act_late])
pFcgLc

# Precision: Probability that the flight was actually late, given that it was flagged
#   p(L | F)
pLgF <- mean(act_late[flag])
pLgF


```


## (3b) Arriving more than three hours late
```{r}


# New Late time
act_late180 <- d$arr_delay > 180
pact_late180 <- mean(act_late180)

# Sensitivity: True positive, probability that the flight was flagged given that it was actually late
#   p(F | L)

pFgL180 <- mean(flag[act_late180])
pFgL180

# Specificity: True negative, probability that the flight wasn't flagged given that it was on time
#   p(Fc | Lc)
pFcgL180c <- mean(!flag[!act_late180])
pFcgL180c

# Precision: Probability that the flight was actually late, given that it was flagged
#   p(L | F)
pL180gF <- mean(act_late180[flag])
pL180gF

```



## (3c) The same precision, by Bayes' rule
```{r}

# p(L180 | F) = [p(F | L180)p(L180)] / p(F)

pL180gF_bayes <- (pFgL180 * pact_late180) / pF
pL180gF_bayes

```


## (3d) For the manager
The flag still captures flights that are late and is good at identifying flights that are at least 30 minutes late, but the number of flights that are 3 hours late or later are less common. There's not enough specificity to rely the flag rule in this context.

# Part 4 · Which distribution, and why not (22 points)

## (4a) Cancellations and scheduled departures

- Cancellations: Daily mean = 266.29; Daily variance = 29999.75. This indicates overdispersion where the rate of cancellations isn't constant or the probability of a cancellation is not independent.

- Daily JFK flights at 5PM: Daily mean = 281.35; Daily variance = 485.64. This includes all 31 days and still shows over-dispersion. When reducing the sample to 28 days - daily mean = 286.68; daily variance = 10.23 (Poisson fails in the opposite direction - underdispersion). Considering that we're looking at a specific time across days, the process that produces a poisson distribution where mean = variance is broken because the rate of the process (flights per day at 5PM) is fixed rather than random or independent.

- Difference in dispersion between 31 days and 28 days for flights at 5PM out of JFK: Not all months have 31 days, so mean and variance when calculated for 31 days is probably biased upward for months that have more data (more days). The difference between mean and variance isn't as large as it is for cancellations because we've taken a smaller sample (JFk at 5PM), rather than all aiports over a 24 hour period of time.

```{r}

# Count flights with missing dep_time for each day
cancellations <- flights %>%
  filter(is.na(dep_time)) %>%
  group_by(day) %>%
  summarise(n = n(), .groups = "drop")
cancellations

cancellations_summary <- cancellations %>%
  summarise(daily_mean = mean(n), daily_var = var(n))
cancellations_summary

# Count flights from JFK with hour == 17 for each day
sched_dept_jfk <- flights %>%
  filter(origin == "JFK", hour == 17) %>%
  group_by(day) %>%
  summarize(n = n(), .groups = "drop")
sched_dept_jfk

# Over-dispersion with all 31 days
sched_dept_jfk_summary <- sched_dept_jfk %>%
  summarise(daily_mean = mean(n), daily_var = var(n))
sched_dept_jfk_summary

# Under-dispersion with fewer days
sched_dept_jfk %>%
  filter(day <= 28) %>%
  summarise(daily_mean = mean(n), daily_var = var(n))

```


## (4b) Late flights per day, JFK to LAX
A binomial distribtuion assumes independence between flights (which is not waht is actually observed), and constant probability across flights (which also isn't observed since flights aren't independent, statistically). Thus, observed variance does not match what we would otherwise expect under those assumptions. Patterns in flight cancellations are correlated at some level - weather, scheduling, etc.

```{r}
# Binomial Distribution
# n = No. of observations (trials)
# p = probability of an event (e.g. late flight)
# mean = np
# variance = np(1-p)

# Assumptions:
# 1) Fixed number of trials, n
# 2) Independence between trials
# 3) Constant probability of event across all trials, p

d_jfk_lax <- d %>% 
  filter(origin == "JFK", dest == "LAX")

# Pooled rate across all days
p_hat_late_jfk_lax <- mean(d_jfk_lax$arr_delay > 15)   # probability of: event = late
n_total_jfk_lax     <- length(d_jfk_lax$arr_delay)      # total no. of flights, all days
p_hat_jfk_lax

# Per day
daily_jfk_lax <- d_jfk_lax %>% 
  group_by(day) %>%
  summarise(total = n(),
            n_late = sum(arr_delay > 15),
            .groups = "drop") %>%
  mutate(theoretical_var = total * p_hat_late_jfk_lax * (1 - p_hat_late_jfk_lax))

observed_var <- var(daily_jfk_lax$n_late)
observed_var

avg_theoretical_var <- mean(daily_jfk_lax$theoretical_var)
avg_theoretical_var


```


## (4c) The 68-95-99.7 rule, twice

```{r}
# 68-98-99.7 for d$arr_delay





# 68-95-99.7 for d_jfk_lax$air_time




```



# Part 5 · < or ≤ (8 points)

## (5a) The daily late count

## (5b) air_time

# Part 6 · The staffing sentence (10 points)

*One paragraph, at most 150 words.*

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
