---
title: "Homework 2"
subtitle: "DATS 6101 · Fall 2026"
author: "Sam Nicholes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r data}
library(nycflights13)
data(flights)
```

<!--
The definitions are on the assignment page, and your numbers are checked against
them, so use them exactly:

  d        rows where both dep_delay and arr_delay are present (327,346 rows)
  A        arr_delay > 15
  D        dep_delay > 15
  a day    the calendar date from year, month, day
  morning  dep_time %/% 100 below 12;  evening: 17 or later

Part 4(a) uses all 336,776 rows instead of d. It says so.

Before you submit: Session > Restart R, then Knit. Every number you state in
prose must come from inline R code, like this: `r nrow(flights)` rows.
-->

# Part 1 · Events are logical vectors (10 points)

## (a) P(A), P(D), P(A and D)

```{r}
d <- flights[complete.cases(flights), ] # data set of complete rows stored as d

A <- d$arr_delay > 15 # flights that arrived >15 min late
head(A, 50) # check vector

D <- d$dep_delay > 15  # flights that departed >15 min late
head(D, 50) # check vector

```



## (b) P(A or D), two ways

```{r}
# P(AuD), probability that flight is late by arrival OR by departure

pA <- mean(d$arr_delay > 15, na.rm = TRUE) # P(late arrival)
pD <- mean(d$dep_delay > 15, na.rm = TRUE) # P(late departure)

# By operator '|'
pAuD_opp <- mean(A | D, na.rm = TRUE)

# By additive rule
pAnD_opp <- mean(A & D, na.rm = TRUE)

pAuD_add <- pA + pD - pAnD_opp

isTRUE(all.equal(pAuD_opp, pAuD_add)) # vectors are the same mathematically

pAuD_opp == pAuD_add # vectors differ by float-point
print(pAuD_opp, digits = 20)
print(pAuD_add, digits = 20)

```



## (c) Neither
The complement of p(AuD) gets us to p(Ac n Dc).
1 - p(AuD) = p(Ac n Dc)

```{r}
# P(flight has on time departure and arrival)
# Complement of p(AuD)
cAuD <- 1 - pAuD_add
cAuD

```




# Part 2 · Independence, checked rather than assumed (18 points)

## (a) The product, the ratio, and P(A | D)

## (b) Late, by origin

*Numbers first, then whether the difference matters.*

## (c) Two flights, same day

# Part 3 · A rule that works, and the same rule that does not (22 points)

## (a) Arriving more than 15 minutes late

## (b) Arriving more than three hours late

## (c) The same precision, by Bayes' rule

## (d) For the manager

# Part 4 · Which distribution, and why not (22 points)

## (a) Cancellations and scheduled departures

## (b) Late flights per day, JFK to LAX

## (c) The 68-95-99.7 rule, twice

# Part 5 · < or ≤ (8 points)

## (a) The daily late count

## (b) air_time

# Part 6 · The staffing sentence (10 points)

*One paragraph, at most 150 words.*

# AI disclosure

<!--
Keep ONE of the two blocks below and delete the other.
"I did not use an AI assistant" is always acceptable and earns full marks.
-->

## Prompt log

## Critique

*150-300 words. Either name something the assistant got wrong and what it would
have cost, or show the prompts you used trying to break it when it held up.
Both earn full marks.*

<!-- OR, replacing both sections above:

I did not use an AI assistant on this assignment.

-->
